---
title: "MachineLearningProject_LCP"
date: "December 16, 2014"
output: html_document
---

Background
========================================================

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 

Data 
=====

The training data for this project are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment. 

What you should submit
=======================

The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases. 

1. Your submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-).
2. You should also apply your machine learning algorithm to the 20 test cases available in the test data above. Please submit your predictions in appropriate format to the programming assignment for automated grading. See the programming assignment for additional details. 

Reproducibility 
================

Due to security concerns with the exchange of R code, your code will not be run during the evaluation by your classmates. Please be sure that if they download the repo, they will be able to view the compiled HTML version of your analysis. 

RESULTS
==========

**1.Load required libraries and data files containing the data:**

```{r}
# install.packages("caret")
# install.packages("randomForest")
# install.packages("rpart")
# install.packages("kernlab")
# install.packages("rattle")
# install.packages("ipred")
library(caret)
library(randomForest)
library(rpart)
library(rpart.plot)
library(rattle)
library(kernlab)
library(ipred)

#Set initial seed for reproducibility
set.seed(54321)

workingDir <- "/Volumes/Data/E-courses/Johns\ Hopkins\ University\ Data\ Science/Practical_Machine_Learning/Project"
#Set the working directory in your computer
setwd(workingDir)

#Check if the files containing the data is already in your working directory. Otherwise, download it, save as a csv file and get the data replacing all missing values as NA
trainingDataURL<-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
trainFilename<-"trainData.csv"
#Check if the file containing the data is already in your working directory. Otherwise, download it, save as a csv file and get the data replacing all missing values as NA
if (file.exists(trainFilename)) {
        trainingData<-read.csv(trainFilename, na.strings=c("NA","#DIV/0!", ""))     
} else {
        download.file(trainingDataURL, destfile = trainFilename, method = "curl")
        trainingData<-read.csv(trainFilename, na.strings=c("NA","#DIV/0!", ""))  
}

testingDataURL<-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
testFilename<-"testData.csv"
if (file.exists(testFilename)) {
        testingData<-read.csv(testFilename, na.strings=c("NA","#DIV/0!", ""))    
} else {
        download.file(testingDataURL, destfile = testFilename, method = "curl")
        testingData<-read.csv(testFilename, na.strings=c("NA","#DIV/0!", ""))
}

```

**2. Preparing the data:**

First of all, I clean the data removing columns only containing missing values (NAs) and unnecessary variables related to data acquisition 
(i.e., [X, user_name, raw_timestamp_part_1, raw_timestamp_part_2, cvtd_timestamp, new_window, num_window]).
Variables with low variance (Near Zero Variance), were also removed.

```{r}
#Delete columns with all missing values
trainingData<-trainingData[,colSums(is.na(trainingData)) == 0]
testingData <-testingData[,colSums(is.na(testingData)) == 0]

#Discard unnecessary variables
trainingData   <-trainingData[,-c(1:7)]
testingData <-testingData[,-c(1:7)]

#Removing near zero covariatees
nearzerovarTr <- nearZeroVar(trainingData, saveMetrics = TRUE)
trainingData <- trainingData[, !nearzerovarTr$nzv]

nearzerovarTe <- nearZeroVar(trainingData, saveMetrics = TRUE)
testingData <- testingData[, !nearzerovarTe$nzv]

dim(trainingData)
dim(testingData)

```

After cleaning the data, **`r length(names(trainingData))`** variables were selected to define the model, i.e., **`r names(trainingData)`**.

Next, I partition the training data set into two subdata sets (i.e., 70% for inTraining, 30% for inTesting) to perform cross-validation. 

```{r}
#Data splitting into two subdata sets

inTrain <- createDataPartition(y=trainingData$classe, p=0.7, list=FALSE)
subTraining <- trainingData[inTrain, ]
subTesting <- trainingData[-inTrain, ]
dim(subTraining)
dim(subTesting)
```

**3. Using different classification algorithms to generate the prediction model:**

I next examined the fit of three different models (Decision Tree, Bagging and Random Forest) to the subTraining data set, and then I tested them on the subTesting data. The best fitting model will be finally used on the test data.

*3.1 Using a decision tree as first prediction model:*

```{r,  fig.width=20, fig.height=15, echo = TRUE}
#Fitting the model
modDT <- rpart(classe ~ ., method = "class", data = subTraining)

#Plotting of the decision tree
rpart.plot(modDT, main="Classification Tree", cex = 1.5, under = TRUE) #faclen=0

#Fancy plotting of the decision tree
fancyRpartPlot(modDT, main="Classification Tree", cex = 1.15, under = TRUE)

#Predicting on the testing subdata set
predictionDT <- predict(modDT, subTesting, type = "class")
cMDT <- confusionMatrix(predictionDT,subTesting$classe)
cMDT
```

*3.2 Using Bagging as second prediction model:*

```{r}
#Fitting the model
modBA <- bagging(classe ~ ., method = "class", data = subTraining)

#Predicting on the testing subdata set and printing the resulting confusion matrix and statistics
predictionBA <- predict(modBA, subTesting, type = "class")
cMBA <- confusionMatrix(predictionBA,subTesting$classe)
cMBA
```

*3.3 Using random forest (improved bagging) as third prediction model:*

```{r}
#Fitting the model
modRF <- randomForest(classe ~ ., method = "class", data = subTraining)
#Predicting on the testing subdata set and printing the resulting confusion matrix and statistics
predictionRF <- predict(modRF, subTesting, type = "class")
cMRF <- confusionMatrix(predictionRF,subTesting$classe)
cMRF
```

**4. Comparing the different models and selecting the best fitting one:**

```{r}
dfAccuracies <- data.frame(cMDT$overall, cMBA$overall, cMRF$overall)
dfAccuracies
```

BY comparing the accuracies of the different models, I select random forest as the one displaying the highest one (Accuracy: **`r dfAccuracies[1,3]`**), with a 95% confidence interval: **`r dfAccuracies[3,3]` - `r dfAccuracies[4,3]`**. As the estimated out-of-sample error rate of the Random Forest trained model (1-Accuracy) is less than 1%, we expect very few errors in predictions. Finally, I display the best fitting final model:

```{r}
#List the variables by relative importance
varImp(modRF)
```



**5. Apply the best fitting model to the test data:**

I then applied the machine Random Forest learning algorithm I built to each of the 20 test cases in the testing data set.
```{r}
# Predict testing results using trained best fitting model
results <- predict(modRF, testingData, type="class")
results

#Save the best fitting model model into a file for subsequent usage
save(modRF, file="modRF.R")

resultsfolder <- "/Volumes/Data/E-courses/Johns\ Hopkins\ University\ Data\ Science/Practical_Machine_Learning/Project/Answers"
setwd(resultsfolder)

pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(results)
```
